Normalization by Evaluation for Call-By-Push-Value and Polarized Lambda Calculus

### 2019-05-10 ppdp19/ppdp19.tex
Added the two references from FSCD review #2. (Warning: appended to auto-ppdp19.bib)
[ ppdp19 ] contributions, acks, conclusion (citing Alechina, Mendler... CSL01)

@@

\section{Conclusion}
\label{sec:concl}

We have defined NbE for CBPV and polarized lambda calculus,
formulated with intrinsically well-typed syntax and presheaf
semantics.  As a side result, we have proven semantically
that the normal forms of both systems are logically complete, \ie,
each derivable judgement $\Gamma \der N$ has a normal derivation.
It remains to show that NbE for these calculi is also computationally
sound and complete, i.e., the computational behavior of term and
normal form should agree, and normalization should decide a suitable
equational theory on terms.

Additionally, a natural question to investigate is whether known CBN
and CBV NbE algorithms can be obtained from our NbE algorithms by
embedding simply-typed lambda calculus into our polarized calculi,
using known CBN and CBV translations.  Further, we would like to study
the NbE algorithm for STLC arising from the optimal translation, i.e.,
the one inserting a minimal amount of $\tThunk$ and $\tComp$ transitions.

\begin{acks}
The first author took initial inspiration from some
  unpublished notes by Thorsten Altenkirch titled
\emph{Another topological completeness proof for intuitionistic logic}
received by email on 16th March 2000.
Thorsten in turn credits his inspiration to an ALF proof by Thierry
Coquand.  Thanks to Hugo Herbelin for feedback and spotting some typos.
\end{acks}

>>>>

To summarize, in this article we make the following contributions:
\begin{itemize}
\item We retell the story of normalization by evaluation (NbE) for intrinsically
  typed lambda terms with weak sum through a generic monadic
  interpreter.  We isolate the services the monad needs to offer in
  order to facilitate reification.

\item We observe that the placement of the monad in the process of the
  type interpretation can be determined by polarization, \ie,
  separating types into positive and negative types.
  The most economic placement of the monad is achieved with deep
  polarization such as in focused calculi.
\item As a first focused calculus, we consider call-by-push-value
  (CBPV) which identifies positive types with value types and negative
  types with computation types.  To our best knowledge, we are the
  first to study NbE for pure, effect-free CBPV.  Therein, we observe
  that the structures of CBPV to control effects naturally structure
  the NbE process as well.
\item Finally, we define NbE for a fully focused calculus, the
  polarized lambda calculus.
  We use a factored presentation of terms and normal forms,
  which uses advanced features of Agda (nested and sized types)
  in our partial formalization.
  % Again, we seem to be the first to define
  % NbE for the polarized lambda calculus, in spite of the substantial
  % proof-theoretic studies it has been subjected to.
\end{itemize}

@@

\section{Conclusion and Further Work}
\label{sec:concl}

We have defined NbE for CBPV and polarized lambda calculus,
formulated with intrinsically well-typed syntax and presheaf
semantics.  At the heart of our development stands the notion of a
cover monad, as alternative to the more common sheaf semantics, to
handle sum types.

As a side result, we have proven semantically
that the normal forms of both systems are logically complete, \ie,
each derivable judgement $\Gamma \der N$ has a normal derivation.
It remains to show that NbE for these calculi is also computationally
sound and complete, i.e., the computational behavior of term and
normal form should agree, and normalization should decide a suitable
equational theory on terms.  To this end, we may switch from the
category of presheaves to the category of Kripke predicates, as
investigated through an Agda formalization.%
\footnote{\url{https://andreasabel.github.io/ipl/html/NbeModel.html}}

Additionally, a natural question to investigate is whether known CBN
and CBV NbE algorithms can be obtained from our NbE algorithms by
embedding simply-typed lambda calculus into our polarized calculi,
using known CBN and CBV translations.  Further, we would like to study
the NbE algorithm for STLC arising from the optimal translation, i.e.,
the one inserting a minimal amount of $\tThunk$ and $\tComp$ transitions.
Finally, we might revisit our interpretation of $\tThunk$ as a monoidal
comonad as suggested by the categorical models of modal logic S4
\cite{alechinaMendlerDePaivaRitter:csl01}.

\begin{acks}
The first author took initial inspiration from some
  unpublished notes by Thorsten Altenkirch titled
\emph{Another topological completeness proof for intuitionistic logic}
received by email on 16th March 2000.
Thorsten in turn credits his inspiration to an ALF proof by Thierry
Coquand.  Thanks to Hugo Herbelin for encouraging feedback and spotting some
typos.  Thanks to the anonymous referees of a previous version of this
paper for the valuable feedback and remarks that helped improving the
text of this article.

The first author acknowledges financial support from
VR Grant 2014-04864 \emph{Termination
      Certificates for Dependently-Typed Programs and Proofs via
      Refinement Types}
and the EU Cost Action CA15123 EUTYPES \emph{Types for Programming and Verification}.
\end{acks}
